Interpretable Risk Minimization -- Preliminaries

### 2024-08-08 paper/main.tex
added theory for multiclass classification, fixed checkpoint recovery bug
committed from 30,000 ft :)

In this setup, each class has it's own set of weights that feedforwards independently, but trains collaboratively. It assigns a score for it's specific value, which is independent since one set of weights does not affect the other over a forward pass. If we take as example logits $[1,2]$ and $[-2,-1]$, we see that $\sigma([1,2]) = \sigma([-2,1]) = [0.2689, 0.7311]$. The prediction probabilities remain the same, despite different values for the same score $s_c$. Therefore, comparing \textit{absolute score} is a misrepresentation of the internal model state.

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c + b_c - s_{c'} - b_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} + b_c - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} - b_{c'}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values, which is an activation map instead. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect. Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}

@@

The distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present.

Segmentations maps provide this exact information. It describes the region over which the object is present. We therefore define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \\
	\mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{y}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{y'}^{\text{HiResCAM}} \\
	\mathcal{\bm{A}}^{\text{background}}_{i,j} := \begin{cases}
		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\mathcal{\bm{A}}^{\text{contrastive}}_{i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := P \cdot \log \frac{P'}{Q} \text{ where }
	P'_{i,j} := \begin{cases}
		1 & P_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{D_1,D_2}_{i,j} \mathcal{\bm{A}}^{\text{foreground}}_{i,j}
\end{gather}

@@

$D_{KL}$ is modified such that the likelihood comparator sets point probability to 1 if it is part of the foreground. While this violates the probability distribution definition, it motivates the optimizer to maximize the foreground activation. During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification negates.

@@

The Oxford-IIIT Pets dataset contains a classification goal alongside high-quality segmentation maps, and the binary classification case (image containing a dog or cat) was trained on. The architecture used for training was ResNet-50, initialized with random parameters and two key modifications:

@@

The model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits. The total time for training was $\approx$1hr.

@@

@@

@@

\paragraph{Extending to the Multiclass Setting} Currently, the implementation computes contrastive activation maps from the target to each other class. However, the best approach to optimization needs to be tested. The likely approach would be to compute the mean error over each contrastive activation map.

>>>>

In this setup, each class has it's own set of weights that feedforwards independently, but trains collaboratively. It assigns a score for it's specific value, which is independent since one set of weights does not affect the other over a forward pass. If we take as example logits $[1,2]$ and $[-2,-1]$, we see that $\sigma([1,2]) = \sigma([-2,1]) = [0.2689, 0.7311]$. The prediction probabilities remain the same, despite different values for the same score $s_c$. Therefore, comparing \textit{absolute score} is a misrepresentation of the internal model state. Instead, the objective is to maximize the \textbf{difference between class logits}.

@@

We therefore derive an optimization objective that attempts to maximize the contrast between the two activations maps. For the binary classification case, $(c, c')$ reflects that correct and incorrect class respectively.
\begin{gather}
	\max_{\theta} s_c + b_c - s_{c'} - b_{c'} \equiv \max_{\theta} \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} + b_c - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} - b_{c'}
\end{gather}
Ignoring the bias terms, our new optimization objective is:
\begin{gather}
	\max_{\theta} \mathcal{\bm{A}}^{\text{contrastive}} := \tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}
\end{gather}
This is equivalent to the original ERM optimization objective, with the difference being the resultant values preserving spatial information as an activation map instead of flattening the feature map. If $\tilde{\mathcal{\bm{A}}}_{c}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}} > 0$, the classification is correct -- if not, the classification is incorrect.

For the multiclass case, our setup now considers a set of classes $c$; $c_t$ is our target class, and $c_{t'}$ consists of a sampled non-target class. The definition for multiclass contrastive activation maps now results in a set of activation maps, defined as follows:
\begin{gather}
	\mathcal{\bm{A}}^{\text{contrastive}} := \{\tilde{\mathcal{\bm{A}}}_{c_t}^{\text{HiResCAM}} - \tilde{\mathcal{\bm{A}}}_{c'}^{\text{HiResCAM}}\}^{|c|-1}_{c_{t'} \in c - c_t}
\end{gather}
This is a generalization of the binary case, as the same formula applied to a binary classification challenge returns a set of contrastive maps containing a single element.

To ensure that the model accurately predicts the target class, each contrastive comparison should yield a positive result. Therefore, the multiclass optimization objective under this paradigm is as follows:
\begin{gather}
	\max_{\theta} \tilde{\mathcal{\bm{A}}}_{i}^{\text{contrastive}}\ \forall i \in \mathbb{N}(|c| - 1)
\end{gather}

@@

As discussed in the previous section, the distinctive feature of the contrastive optimization objective over standard risk is that this preserves spatial information. We know at a pointwise level how each feature contributes to the final prediction.

We can use this to constrain the way optimization occurs -- namely, we only want the model to increase the likelihood of an object belonging to a certain class by actually looking at the features of the class, and not features of spurious correlation like the environment in which the object is present. The model through the process of training should learn to distinguish and subsequently ignore the background in determining the final prediction.

Segmentations maps describes the region over which the object is present. We can leverage this information to define a constrastive cost function as follows:
\begin{gather}
	\{(x, (h, y))\}^N \in \mathcal{D} \text{ where } h_{i,j} := \begin{cases}
		1 & \text{pixel contains subject} \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{background}}_{c,i,j} := \begin{cases}
		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & h_{i,j} = 0 \\
		0 & \text{otherwise}
	\end{cases} \\
	\mathcal{\bm{A}}^{\text{foreground}}_{c,i,j} := \begin{cases}
		0 & h_{i,j} = 0 \\
		\mathcal{\bm{A}}^{\text{contrastive}}_{c,i,j} & \text{otherwise}
	\end{cases} \\
	D_{KL}(P || Q) := \frac{1}{|c|-1}\sum^{|c|-1}_c P \cdot \log \frac{P'}{Q_c} \text{ where } 
	P'_{i,j} := \begin{cases}
		1 & P_{i,j} \neq 0 \\
		P_{i,j} & \text{otherwise}
	\end{cases} \label{d_kl} \\
	\mathcal{\bm{J}}(\theta, \mathcal{D}) := D_{KL}(\sigma_{\text{softmax}}(\lambda_1 \mathcal{\bm{A}}^{\text{contrastive}})\ ||\ \sigma_{\text{softmax}}(\lambda_2 \cdot h)) + \frac{\lambda_3}{D_1 D_2} ||\mathcal{\bm{A}}^{\text{background}}||^2_2 - \frac{1}{D_1 D_2} \sum^{|c|-1,D_1,D_2}_{c,i,j} \mathcal{\bm{A}}^{\text{foreground}}_{c,i,j}
\end{gather}

@@

$D_{KL}$ is modified per (10), such that the likelihood comparator sets point probability to 1 if it is part of the foreground. While this violates the probability distribution definition, it motivates the optimizer to maximize the foreground activation. During training, observing pointwise divergence revealed that the model was penalized for `over-activating' the foreground, which this modification corrects. In addition, it also averages the divergence from the target segmenetation map to each contrastive activation map.

@@

The Oxford-IIIT Pets dataset is an image dataset with a classification objective alongside high-quality segmentation maps. We train the model on the binary classification case (image containing a dog or cat) as well as the fine-grained, breed specific 37 class split. The architecture used for training was ResNet-50, initialized with random parameters and three key modifications:

@@

@@

\subsection{Binary Classification}

For binary classification, the model was trained for $150$ epochs with the Adam optimizer with a learning rate of $10^{-3}$, a batch size of $32$ on a 32GB V100 on a randomly shuffled dataset using the original data splits. The total time for training was $\approx$1hr.

@@

\subsection{Multiclass Classification}

%TODO: update results
\begin{table}[h]
	\centering
	\begin{tabular}{c|ccc}
		\toprule
		\textbf{Method}  & \textbf{Cross Entropy Loss}  & \textbf{Training Accuracy}   & \textbf{Validation Accuracy} \\
		\midrule
		Cross-Entropy & 0.000 & 3.0\% & 3.0\% \\
		Interpretable (Ours) & \bf 0.000 & \bf 92.0\% & \bf 50.0\% \\
		\bottomrule
	\end{tabular}
\end{table}

@@

\paragraph{Understanding the Generalization Gap} While the model for the multiclass paradigm is able to fit the training dataset excellently -- achieving $\approx 92\%$ training accuracy -- the validation accuracy is significantly behind at $\approx 50\%$. This could imply memorization of samples, which was something that the approach was intended to mitigate. Further investigation could help support or disprove this result.
